Real-time modelling of a public transport network to improve arrival-time predictions

### 2018-02-05 realtime-filter/realtime-filter.Rnw
matrix notation and new symbols

The main goal of our work is to estimate arrival times of buses,
which requires knowledge of how long it takes a bus to
travel along roads within the network.
Thus, we need to estimate the \emph{road network state} at time $c$,
\begin{equation}
  \label{eq:kf_state}
  \bbeta_c = [\beta^1_c, \cdots, \beta^M_c]^T
\end{equation}
where $\beta^j_c$ is the travel time along road segment $j$.
The basic idea is to update this state every time a bus travels
along a segment of road.
This way, subsequent buses can use the most recent travel time
information to estimate how long it will take them to reach 
upcoming stops.

Due to the simplicity of the model (a single parameter, travel time),
we are using an \gls{ekf} algorithm; 
we are modeling each segment independently, although in the future
we plan to take into account adjacent segments.
However, since we cannot expect regular observations,
we need the state to "fall back" on our prior estimate of travel time%
---based on historical data---over time.

The first stage is to derive the transition function $f_c$ that
trends towards the prior mean, $\mu_j(t)$, and variance, $\omega_j(t)^2$,
of travel times along segment $j$ at time $t$ (based on historical data).
For now, we will be using $\mu_j = \mu_j(t)$ and $\omega_j = \omega_j(t)\ \forall t$.
We also need to allow $f_c$ to cope with irregular update intervals 
($\Delta_c = $ time since segment was last updated).

We first define a one-second transition function $f$ such that the state mean
converges towards the prior mean,
\begin{equation}
  \label{eq:kf_trans1}
  \beta_{c} = f(\beta_{c-1}, \mu, \lambda) = \beta_{c-1} + \lambda(\mu - \beta_{c-1}),
\end{equation}
where $\lambda \in (0, 1)$ is a tuning parameter that determines how quickly
the process converges to the prior.
We next define $f_n$ as a recusive composition of $n$ transitions,
\begin{equation}
  \label{eq:kf_transf}
f_n(\beta, \mu, \lambda) =
(f^1\circ \cdots \circ f^n) (\beta, \mu, \lambda),\quad
\text{where } f^j \equiv f\ \forall j.
\end{equation}

For the \gls{ekf}, we need the derivative of $f_n$, which we get via the chain rule.
To simplify notation, we're dropping the $\mu$ and $\beta$ parameters since they constant.
\begin{align}
  \label{eq:kf_deriv}
  F_n = f_n'(\beta)\nonumber
    &= f'[(f^1\circ \cdots \circ f^{n-1})(\beta)]
      (f^1\circ \cdots \circ f^{n-1})'(\beta) \nonumber \\
    &= f'[f_{n-1}(\beta)] F_{n-1} \nonumber \\
  \intertext{which, by recursion, we get}
    &= f'[f_{n-1}(\beta)] f'[f_{n-2}(\beta)] \cdots f'[f_1(\beta)] f'(\beta) \nonumber \\
    &= f'(\beta) \prod_{i=1}^{n-1} f'[f_i(\beta)] \nonumber \\
  \intertext{Next we can calculate 
  $f'(\beta, \mu, \lambda) = \frac{\partial}{\partial \beta} \beta + \lambda(\mu - \beta) = 1 - \lambda$,
  so the derivative becomes}
    F_n &= (1-\lambda) \prod_{i=1}^n (1 - \lambda) = (1-\lambda)^n
\end{align}

That ensure that as $n$ goes to infinity, $f_n(\beta, \mu, \lambda) \to \mu$.
To do the same with the variance, we need to find a value for $Q$,
the variance of system noise, which is involved in the prediction of the state variance, $\Sigma$:
\begin{align}
  \label{eq:kf_varpred}
  \Sigma_{c|c-1} &= F_n^2\Sigma_{c-1|c-1} + Q \nonumber \\
  \intertext{At equilibrium, we want $\Sigma_{c|c-1} = \Sigma_{c-1|c-1} = \omega$, so}
  \omega &= F_n^2 \omega + Q = (1-\lambda)^{2n} \omega + Q \nonumber \\
  \intertext{which, solving for $Q$, gives us}
  Q &= \omega(1 - (1 - \lambda)^{2n})
\end{align}

The last step is to choose a value of $\lambda$.
However, any fixed value will cause the system to converge quickly when the difference is large;
instead, we want the convergence to be a function of the ratio of information held
in the prior and posterior states. 
For example, if we have a lot of buses traveling at the same speed along a segment, 
we will have a lot of information and want the state estimate to remain there
longer that if only one bus went through. 
So we will use $\lambda = \phi \frac{\Sigma_{c-1|c-1}}{\Sigma_{c-1|c-1} + \omega}$.

>>>>

The main goal of our work is to estimate arrival times of buses,
which requires knowledge of how long it takes a bus to
travel along roads within the network.
Thus, we need to estimate the \emph{road network state} at time $c$,
\begin{equation}
  \label{eq:kf_state}
  \bbeta_c = [\beta^1_c, \cdots, \beta^M_c]^T
\end{equation}
where $\beta^m_c$ is the travel time along road segment $m$.
The basic idea is to update this state every time a bus travels
along a segment of road.
This way, subsequent buses can use the most recent travel time
information to estimate how long it will take them to reach 
upcoming stops.

We are using an \gls{ekf} algorithm to model the network state,
treating each segment independently for now, although in the future
we plan to take into account adjacent segments
(see section~\ref{sec:discussion}).
However, since we cannot expect regular observations,
we need the state to ``fall back'' on a prior travel time distribution%
---based on historical data---over time.

The model formulation for the network state is therefore
\begin{align}
  \label{eq:kf_model}
  \bbeta_c &= \mat{F}_c(\bbeta_{c-1}, \bnu_c, \bpsi_c) + \bu_c \\
  \bB_c &= \bbeta_c + \be_c,
\end{align}
where $\mat{F}_c$ is the transition function at time $c$ that
trends towards the prior mean, $\bnu_c$, at some rate determined
by the turning parameter $\bpsi_c$.
Note that all of these parameters are vectors with $M$-components.

The first stage is to derive the transition function $\mat{F}_c$ that
trends towards the prior mean, $\bnu_c$, based on historical data.
We also need to allow $\mat{F}_c$ to cope with irregular update intervals 
($\Delta_c = t_c - t_{c-1} = $ time since segment was last updated).

We first define a one-second transition function $\mat{F}$ such that the state mean
converges towards the prior mean,
\begin{equation}
  \label{eq:kf_trans1}
  \bbeta_{c} = \mat{F}_c(\bbeta_{c-1}, \bnu, \bpsi) = \bbeta_{c-1} + \bpsi_c\circ(\bnu - \bbeta_{c-1}),
\end{equation}
where $\psi^m_c \in (0, 1)\ \forall m$ is a tuning parameter that determines how quickly
the process converges to the prior.
We next define $\mat{F}_n$ as a recusive composition of $n$ transitions,
\begin{equation}
  \label{eq:kf_transf}
\mat{F}_n(\bbeta, \bnu, \bpsi) =
(\mat{F}^1\circ \cdots \circ \mat{F}^n) (\bbeta, \bnu, \bpsi),\quad
\text{where } \mat{F}^j \equiv \mat{F}\ \forall j.
\end{equation}

For the \gls{ekf}, we need the derivative of $\mat{F}_n$, which we get via the chain rule.
To simplify notation, we're dropping the $\bnu$ and $\bpsi$ parameters since they're constant,
and also only considering each segment independently, so
\begin{align}
  \label{eq:kf_deriv}
  \F_n' (\bbeta) &= 
    \begin{bmatrix}
      F_n' (\beta^1) \\ \vdots \\ F_n' (\beta^M)
    \end{bmatrix}\nonumber \\
  \text{and } F_n'(\beta)\nonumber
    &= F'[(F^1\circ \cdots \circ F^{n-1})(\beta)]
      (F^1\circ \cdots \circ F^{n-1})'(\beta) \nonumber \\
    &= F'[\F_{n-1}(\beta)] F'_{n-1}(\beta) \nonumber \\
  \intertext{which, by recursion, we get}
    &= F'[F_{n-1}(\beta)] F'[F_{n-2}(\beta)] \cdots F'[F_1(\beta)] F'(\beta) \nonumber \\
    &= F'(\beta) \prod_{i=1}^{n-1} F'[F_i(\beta)] \nonumber \\
  \intertext{Next we can calculate 
  $F'(\beta, \nu, \psi) = \frac{\partial}{\partial \beta} \left[\beta + \psi(\nu - \beta)\right] = 1 - \psi$,
  so the derivative becomes}
    F'_n(\beta) &= (1-\psi) \prod_{i=1}^n (1 - \psi) = (1-\psi)^n
\end{align}

That ensure that as $n$ goes to infinity, $\F_n(\bbeta, \bnu, \bpsi) \to \bnu$.
To do the same with the variance, we need to find a value for $\bQ$,
the variance of system noise, which is involved in the prediction of the state variance, $\bP$:
\begin{align}
  \label{eq:kf_varpred}
  \bP_{c|c-1} &= \F_n \bP_{c-1|c-1}\F_n^T + \bQ \nonumber \\
  \intertext{At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
  which we can also compute independently for each segment (since the off-diagonals will be 0):}
  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m} = (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
  \intertext{which, solving for $Q$, gives us}
  Q_{m,m} &= \xi^m(1 - (1 - \psi^m)^{2n})
\end{align}

The last step is to choose a value of $\bpsi$.
However, any fixed value will cause the system to converge quickly when the difference is large;
instead, we want the convergence to be a function of the ratio of information held
in the prior and posterior states. 
For example, if we have a lot of buses traveling at the same speed along a segment, 
we will have a lot of information and want the state estimate to remain there
longer that if only one bus went through. 
So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$.
